The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-08-06 script/en/chapters/numbertheory.tex
new float cryptoprocedure for crypto algorithms and procedures (chapter 5); fixed links to un-numbered chapters with phantomsection
resolved more pdfTeX warnings by adding a chapter specific prefix "xy:" to all cite and bibitem in chapters with bibliography

Further mathematical disciplines currently used in cryptography include
(see \cite[p. 2]{Bauer1995}, \cite[p. 3]{Bauer2000}) :
\begin{itemize}
    \item Group theory\index{Group}
    \item Combination theory
    \item Complexity theory
    \item Ergodic theory
    \item Information theory.
\end{itemize}

Number theory or arithmetic (the emphasis here is more on the aspect of 
performing calculations with numbers) was established 
by Carl Friedrich Gauss\footnote{%
  Carl Friedrich Gauss, German mathematician and astronomer,
  Apr 30, 1777$-$Feb 23, 1855.
}
\index{Gauss, Carl Friedrich} as a special mathematical discipline. Its 
elementary features include the greatest common divisor\footnote{This 
article deals with the gcd (greatest common 
divisor)\index{Gcd} in \hyperlink{NumberTheory_Appendix_A}{Appendix A of this chapter}.}
(gcd), congruence (remainder classes), factorization, the Euler-Fermat theorem
and primitive roots. However, the most important 
aspect is prime numbers and their multiplicative operation.

@@

We now know that patterns from number theory can be found everywhere in nature. 
For example, the ratio of 
%laevorotary and dextrorotary 
rotating counterclockwise and rotating clockwise 
spirals in a sunflower is equal to two consecutive 
Fibonacci\index{Fibonacci} numbers\footnote{%
The sequence of Fibonacci numbers $(a_i)_{i \in \mathbb{N}}$ is defined 
by the ``recursive'' rule $a_1 := a_2 := 1$ and for all numbers  $n=1,2,3,\cdots$ 
we define $a_{n+2} := a_{n+1}+a_n$.  
This historical sequence can be found in many interesting forms in nature 
(for example, see \cite[p. 290 ff]{Graham1994}\index{Graham 1994} or the 
website of \hyperlink{knott}{Ron Knott}\index{Knott, Ron}, which is devoted 
to Fibonacci\index{Fibonacci} numbers). A lot is known about the Fibonacci 
sequence and it is used today as an important tool in mathematics.}, for 
example $21 : 34$.

@@

Applications in (computer) security now use cryptography because this mathematical discipline is simply better 
and easier to prove than all other ''creative'' substitution procedures that have been developed over the course of 
time and better than all sophisticated physical methods such as those used to print bank notes \cite[p. 4]{Beutelspacher1996}.

@@

Number theory has always been a source of fascination -- for both amateurs and
professional mathematicians. In contrast to other areas of mathematics, many
of the problems and theorems in number theory can be understood by non-experts.
On the other hand, the solutions to these problems or the prove to the theorems
often resisted to the mathematicians for a very long time.
It is therefore one thing to pose good questions but quite another matter to
find the answer. One example of this is what is known as Fermat's Last (or
large) theorem\footnote{One of the things we learn in mathematics at school
   is Pythagoras' theorem, which states the following for a right-angle
   triangle: $a^2 + b^2 = c^2$, where $a$ and $b$ are the lengths of the sides
   containing the right angle and $c$ is the length of the hypotenuse.
   Fermat famously proposed that $a^n + b^n \not= c^n$
   for $a,b,c \in \mathbb{N}$ and whole-number exponents $n > 2$.
   Unfortunately, the letter in which Fermat made the claim did not have
   enough space for him to prove it. The theorem was not proven until over
   300 years later \cite[p. 433-551]{Wiles1994}\index{Wiles, Andrew}. }.

@@

Although it may seem difficult to calculate the table of multiplicative inverses for large moduli (this only applies 
to the areas of the table shaded dark grey), we can use Fermat's Little Theorem\index{Fermat!little theorem}
to create a simple algorithm for this \cite[p. 80]{Pfleeger1997}. Quicker algorithms are described, for instance, in 
\cite{Knuth1998}\footnote{Using Euclid's extended theorem\index{Euclidean algorithm!extended}  (extended gcd), we can calculate the 
multiplicative inverse and determine whether numbers have an inverse (see \hyperlink{NumberTheory_Appendix_A}{appendix A of this chapter}). Alternatively, we can also use the primitive roots.}.

@@

The situation is similar if 
$b$ is the unknown variable and we know the values of $a$ and $c$: \\
If we know the Euler function \index{Euler!(phi) function} of $m, J(m)$, then we can easily\footnote{See \hyperlink{NumberTheory_Appendix_A}{Appendix A of this chapter}:
the greatest common divisor (gcd) of whole numbers.}\index{Gcd} calculate $d$ with $c*d \equiv 1 {\rm ~(mod~} J(m))$ and 
use theorem~\ref{thm-zth-pot} to obtain:
$$
   a^d \equiv (b^c)^d \equiv b^{c*d} \equiv b^{c*d~(mod~J(m))} \equiv b^1 \equiv b {\rm ~(mod~} m)
$$
the {\em $c$-th root} $b$ of $a$. \par

@@

\begin{theorem}\label{thm-zth-ordp}\footnote{%
    For prime moduli $p$ all $a$ with $0 < a < p$ are of order $J(p) = p - 1$.
    Compare table~\ref{modexp45} for an example. In this case $a^i (\mbox{mod }n)$ goes
    through all the values $1,\dots,p-1$. Exhausting all possible values of
    the set is an important cryptographic proposition (compare theorem
    \ref{thm-zth-exhperm}). This determines a permutation\index{Permutation}
    $\pi(p-1)$.}$^,$\footnote{%
    Table~\ref{modexp46} demonstrates that for composite moduli $n$ not all $a$ are of
    maximal order $J(n)$. In this example only $5,7,11,15,17,19\mbox{ and }21$
    are of order 22.
  } Given a modulus $n$ and a number $a$, relative prime to $n$ the
  following holds: \\
  The set $\{ a^i\; (\mbox{mod }n) |\; i = 1,\dots,J(n)\}$ equals the
  multiplicative group $Z_n^*$ if and only if $ord_n(a) = J(n)$.
\end{theorem}
The multiplicative group $Z_n^*$ contains all values from $1$ to $n-1$.

@@

% --------------------------------------------------------------------------
\hypertarget{RSA}{}
\subsection{How the RSA procedure works}
\label{RSA}
The individual steps for implementing the \index{RSA} RSA procedure can be described as follows (see \cite[p. 213 ff]{Eckert2003}\index{Eckert 2003} and 
\cite[p. 338 ff]{Sedgewick1990}\index{Sedgewick 1990}). 
Steps 1 to 3 constitute key generation, steps 4 and 5 are the encryption, and steps 
6 and 7 are the decryption:
\begin{itemize}
\item[{\bf 1.}] Select two distinct random prime numbers\footnote{
Compaq introduced the so-called multi-prime method with high marketing 
effort in 2000.
$n$ was the product of two big and one relative small prime: $n=o*p*q.$ 
With theorem~\ref{J_of_p1..pk} we get:
$J(n)= (o-1)*(p-1)*(q-1). $ This method did not assert itself yet.\\
One reason probably is, that Compaq claimed a patent \index{Patent} on it. 
Generally there is less understanding in Europe and with the Open Source
Initiative\index{Open Source}, that one can claim patents on algorithms.
But there is really no understanding outside the U.S., that one can get
a patent for a special case (3 factors) of an algorithm (RSA), although
the patent for the general case was almost expired.}$^,$\footnote{If the
two primes $p$ and $q$ are equal then $ (m^{e})^{d} \equiv m $ mod $n$ is
not true for all $ m<n$ (although $ e*d \equiv 1$ mod $J(n)$ is fulfilled). 
\begin{example}{:}\\
If $ n=5^2$ then according to theorem~\ref{thm-zth-phinum} it 
is $J(n)=5*4=20, ~ e=3, ~ d=7, ~e*d=21\equiv 1$ mod $J(n).$
\end{example}
But it is $ (5^3)^7\equiv 0 $ mod $ 25$. } $p$ and $q$ and calculate 
$n = p*q$\footnote{The GISA\index{GISA} (German Information Security Agency) 
recommends, to choose the prime factors $p$ and $q$ almost the same,
but not too close:
$$ 0.5 < |\log_2 (p) - \log_2 (q) | <30. $$
They recommend to generate the primes independently and check that the 
restriction is fulfilled (see \cite{GISA2002}).
}.\\
The value $n$ is called the RSA modulus\footnote{
In CrypTool\index{CrypTool} the RSA modulo is denoted with a capital ``$N$'' .}.

@@

\noindent {We will show this in 3 steps (see \cite[p. 131ff]{Beutelspacher1996}).

\noindent {{\bf Step 1}:

@@

\noindent {{\bf Step 2}:

@@

\noindent {{\bf Step 3}:

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{SecurityRSA}{}
\section[Considerations regarding the security of the RSA algorithm]
{Considerations regarding the security of the RSA algorithm\footnotemark}
    \footnotetext{Major
    parts of chapters~\ref{complexity} and \ref{chptSecurityParam} follow
    the article ``Vorz\"uge und Grenzen des RSA-Verfahrens'' written by
    F.~Bourseau, D.~Fox and C.~Thiel \cite{Bourseau2002}.}
    \label{SecurityRSA}

@@

The best factorization method known today is a further development of the
General Number Field Sieve (GNFS) \index{General Number Field Sieve (GNFS)},
which was originally devised to factor only numbers of a special form 
(like Fermat numbers).  The complexity of solving the factorization problem
with the GNFS is asymptotically
$$
O(l) = e^{c \cdot (l \cdot \ln 2)^{1/3} \cdot  (\ln(l \cdot \ln(2))^{2/3} + o(l)}
$$
Please refer to: 
\vspace{-10pt}
\begin{itemize}{}
  \item A. Lenstra, H. Lenstra:  
          {\em The development of the Number Field Sieve} 
          \cite{Lenstra1993}.
  \item   Robert D. Silverman:  
          {\em A Cost-Based Security Analysis of Symmetric and Asymmetric
          Key Lengths} 
          \cite{Silverman2000}.
\end{itemize}

@@

Please refer to: {\em Handbook of Applied Cryptography} \cite{Menezes2001}.

@@

      Please refer to:
      \begin{list}{}{\setlength{\topsep}{-7 pt}}
        \item[] C. Pomerance:  
                {\em The quadratic sieve factoring algorithm} 
                \cite{Pomerance1984}.
      \end{list}
      \vskip +12pt

@@

      Please refer to:
      \begin{list}{}{\setlength{\topsep}{-7 pt}}
        \item[] J.P. Buhler, H.W. Lenstra, C. Pomerance:  
                {\em Factoring integers with the number field sieve} 
                \cite{Buhler1993}.
      \end{list}
      \vskip +12pt
\end{itemize}

@@

A good overview about the current status of publications concerning these attacks
can be found in the diploma thesis of Matthias Schneider \cite{SchneiderM2004}.

@@

An estimation of the evolution of secure key lengths was done by 
Lenstra/Verheul\cite{Lenstra1999} (compare figure~\ref{RSAKeylength}
in chapter ~\ref{ECAlternative}).

Within the article \cite{Bourseau2002} Dirk Fox\footnote{%
His company Secorvo Ltd also delivered a statement on the recommendation for key
length selection published by the GISA\index{GISA} (German Information
Security Agency). Chapter 2.3.1 of this statement contains a competent
and understandable discussion of RSA security (this document
exists -- to my knowledge -- only in German):\\
\href{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}
{\texttt{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}}
}
published his prognosis of an almost linear factorization progression,
if all influencing factors are included: Each year the module length feasible for
factorization increases by 20 bit on average. So his forecast was below the
more optimistic estimations of GISA and NIST.

@@

In his paper ``Circuits for integer factorization: a proposal'' 
(\href{http://cr.yp.to/djb.html}{\texttt{http://cr.yp.to/djb.html}}),
published November 2001, D.~J.\ Bernstein \cite{Bernstein2001} 
addresses the problem of
factorizing large integers. Therefore, his results are of relevance from a
RSA point of view.  As a main result Bernstein claims that the
implementation of the General Number Field Sieve algorithm (GNFS)
 \index{General Number Field Sieve (GNFS)} can be improved to factor, with
the same effort as before, integers with three times more digits.

@@

Arjen Lenstra, Adi Shamir et. al. analyzed the paper of Bernstein
\cite{Lenstra2002}.  In summary they expect a factorization improvement on
how much longer the bit length of the keys could be with a factor of 1.17
(instead of factor 3 as proposed by Bernstein).

@@

\href{http://www.rsasecurity.com/}{RSA Security's} analysis of the Bernstein
paper \cite{RSA Security 2002} from April, 8 2002 also -- as expected --
concludes, that RSA is still not compromised.

@@

In January 2003 Adi Shamir and Eran Tromer from the Weizmann Institute of Science published a preliminary draft called {\em ``Factoring Large Numbers with the TWIRL Device''} raising concerns about the security of key sizes till 1024 bits \cite{Shamir2003}.

@@

A detailed explanation from these two authors also can be found in the
RSA Laboratories CryptoBytes \cite{Shamir2003a}.

The 3-page article in the DuD issue of June 2003 \cite{Weis2003} contains
a very good explanation, how the attack using the Generalized Number Field
Sieve (GNFS) \index{General Number Field Sieve (GNFS)} works and which 
progress is made, to factorize numbers.
At GNFS we can distinguish 2 general steps: 
The sieve step (relation collecting) and the matrix reduction.
Besides the sieve step is highly parallelizable, it dominates the overall
calculation burden. Shamir and Tromer haven't built a TWIRL device yet,
but the estimated costs of 10 till 50 million Euro (in order to factorize
a 1024-bit number) is not prohibitive for secret agencies or big criminal
organizations, because the ``costs for a single espionage satellite is
estimated e.g. to be several billion USD''. The authors therefore
recommend, to get as soon as possible rid of today used sensible RSA, 
Diffie-Hellman or ElGamal keys up to 1024 bit and to use then keys of at
least 2048 bit length.
The planned TCPA/Palladium hardware \index{Palladium} will use 2048-bit
RSA keys!

@@

In August 2002 the three Indian researchers M. Agrawal, N. Kayal and N. Saxena published the paper {\em ``PRIMES in P''} about a new primality testing algorithm called AKS\index{AKS} \cite{Agrawal2002}. 
They discovered a polynomial\index{Polynomial} time deterministic algorithm for determining if a number is prime or not.

@@

Alice and Bob\footnote{Bob\index{Bob} and Alice\index{Alice} are the default
names used for the two authorized participants in a protocol (see
\cite[p. 23]{Schneier1996nt}).
} use a one way function to obtain a key $S$, the session key, for subsequent
correspondence. This is then a secret that is only known to the two of them.
Alice selects a random number $a$ and keeps it secret. She applies a one way
function to $a$ to calculate the number $A = g^a$ and sends it to Bob. He does
the same, by selecting a secret random number $b$, calculating $B = g^b$ and
sending it to Alice. The number $g$ is random and can be publicly known.
Alice applies the one way function together with her secret number $a$ to
$B$, while Bob does the same with his secret number $b$ and the received
number $A$.

@@

\noindent {\bf Key generation (steps 1 to 3)}:\\
{\bf 1.} $p=47, q=79$ $( n= 3713; ~J(n) = (p-1)*(q-1)=3588).$ \\
{\bf 2.} $e = 37$ (should\footnote{%
                See footnote~\ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $79$ and $3587$ and must be relatively prime
                to $3588$). \\
{\bf 3.} $d=97$ (since $e*d=1{\rm ~mod~}J(n); 37*97 \equiv 3589 \equiv 1{\rm ~(mod~}3588) \;$)\footnote{How
to compute $d = 97$ using the {\em extended} gcd algorithm is shown in \hyperlink{NumberTheory_Appendix_A}{appendix A of this chapter}}. \\

@@

\noindent The idea for this task\footnote{Using CrypTool\index{CrypTool} you can 
solve this via the menu path
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.} 
is taken from the example in \cite[p. 271]{Eckert2003}\index{Eckert 2003}.

@@

The task is taken from \cite[Exercise 4.6]{Stinson1995}\index{Stinson 1995}:
The pure solution has been published by Prof. Stinson at 
\href{http://www.cacr.math.uwaterloo.ca/~dstinson/solns.html}{\texttt{http://www.cacr.math.uwaterloo.ca/\~{}dstinson/solns.html}}.\footnote{or
\href{http://bibd.unl/~stinson/solns.html}{\texttt{http://bibd.unl/\~{}stinson/solns.html}}.}

@@

The following task is a corrected version from the book written by Prof. Yan 
\cite[Example 3.3.7, p. 318]{Yan2000}\index{Yan 2000}.
However, it is not the result that is important here but rather the individual steps of the solution, that is, 
the explanation of the cryptanalysis\index{Cryptanalysis}\footnote{The method of solving the problem is outlined in the scenario of 
the online help to CrypTool\index{CrypTool} and in the CrypTool presentation. If anyone sends us a well prepared exact 
method of solving the problem, we would be pleased to include it in the documentation.}.

@@

\bibitem[Agrawal2002]{Agrawal2002}  \index{Agrawal 2002} 
    M. Agrawal, N. Kayal, N. Saxena, \\
    {\em PRIMES in P}, August 2002, revised paper: \\
       \href{http://www.cse.iitk.ac.in/~manindra/algebra/primality_v6.pdf}
    {\texttt{http://www.cse.iitk.ac.in/~manindra/algebra/primality\_v6.pdf}} \\
    See also the website "The AKS "PRIMES in P" Algorithm Resource":
       \href{http://fatphil.org/maths/AKS/}
    {\texttt{http://fatphil.org/maths/AKS/}}

>>>>

Further mathematical disciplines currently used in cryptography include
(see \cite[p. 2]{nt:Bauer1995}, \cite[p. 3]{nt:Bauer2000}) :
\begin{itemize}
    \item Group theory\index{Group}
    \item Combination theory
    \item Complexity theory
    \item Ergodic theory
    \item Information theory.
\end{itemize}

Number theory or arithmetic (the emphasis here is more on the aspect of 
performing calculations with numbers) was established 
by Carl Friedrich Gauss\footnote{%
  Carl Friedrich Gauss, German mathematician and astronomer,
  Apr 30, 1777$-$Feb 23, 1855.
}
\index{Gauss, Carl Friedrich} as a special mathematical discipline. Its 
elementary features include the greatest common divisor\footnote{This 
article deals with the gcd (greatest common 
divisor)\index{Gcd} in \hyperlink{NumberTheory_Appendix_GCD}{Appendix A of this chapter}.}
(gcd), congruence (remainder classes), factorization, the Euler-Fermat theorem
and primitive roots. However, the most important 
aspect is prime numbers and their multiplicative operation.

@@

We now know that patterns from number theory can be found everywhere in nature. 
For example, the ratio of 
%laevorotary and dextrorotary 
rotating counterclockwise and rotating clockwise 
spirals in a sunflower is equal to two consecutive 
Fibonacci\index{Fibonacci} numbers\footnote{%
The sequence of Fibonacci numbers $(a_i)_{i \in \mathbb{N}}$ is defined 
by the ``recursive'' rule $a_1 := a_2 := 1$ and for all numbers  $n=1,2,3,\cdots$ 
we define $a_{n+2} := a_{n+1}+a_n$.  
This historical sequence can be found in many interesting forms in nature 
(for example, see \cite[p. 290 ff]{nt:Graham1994}\index{Graham 1994} or the 
website of \hyperlink{knott}{Ron Knott}\index{Knott, Ron}, which is devoted 
to Fibonacci\index{Fibonacci} numbers). A lot is known about the Fibonacci 
sequence and it is used today as an important tool in mathematics.}, for 
example $21 : 34$.

@@

Applications in (computer) security now use cryptography because this mathematical discipline is simply better 
and easier to prove than all other ''creative'' substitution procedures that have been developed over the course of 
time and better than all sophisticated physical methods such as those used to print bank notes \cite[p. 4]{nt:Beutelspacher1996}.

@@

Number theory has always been a source of fascination -- for both amateurs and
professional mathematicians. In contrast to other areas of mathematics, many
of the problems and theorems in number theory can be understood by non-experts.
On the other hand, the solutions to these problems or the prove to the theorems
often resisted to the mathematicians for a very long time.
It is therefore one thing to pose good questions but quite another matter to
find the answer. One example of this is what is known as Fermat's Last (or
large) theorem\footnote{One of the things we learn in mathematics at school
   is Pythagoras' theorem, which states the following for a right-angle
   triangle: $a^2 + b^2 = c^2$, where $a$ and $b$ are the lengths of the sides
   containing the right angle and $c$ is the length of the hypotenuse.
   Fermat famously proposed that $a^n + b^n \not= c^n$
   for $a,b,c \in \mathbb{N}$ and whole-number exponents $n > 2$.
   Unfortunately, the letter in which Fermat made the claim did not have
   enough space for him to prove it. The theorem was not proven until over
   300 years later \cite[p. 433-551]{nt:Wiles1994}\index{Wiles, Andrew}. }.

@@

Although it may seem difficult to calculate the table of multiplicative inverses for large moduli (this only applies 
to the areas of the table shaded dark grey), we can use Fermat's Little Theorem\index{Fermat!little theorem}
to create a simple algorithm for this \cite[p. 80]{nt:Pfleeger1997}. Quicker algorithms are described, for instance, in 
\cite{nt:Knuth1998}\footnote{Using Euclid's extended theorem\index{Euclidean algorithm!extended}  (extended gcd), we can calculate the 
multiplicative inverse and determine whether numbers have an inverse (see \hyperlink{NumberTheory_Appendix_GCD}{appendix A of this chapter}). Alternatively, we can also use the primitive roots.}.

@@

The situation is similar if 
$b$ is the unknown variable and we know the values of $a$ and $c$: \\
If we know the Euler function \index{Euler!(phi) function} of $m, J(m)$, then we can easily\footnote{See \hyperlink{NumberTheory_Appendix_GCD}{Appendix A of this chapter}:
the greatest common divisor (gcd) of whole numbers.}\index{Gcd} calculate $d$ with $c*d \equiv 1 {\rm ~(mod~} J(m))$ and 
use theorem~\ref{thm-zth-pot} to obtain:
$$
   a^d \equiv (b^c)^d \equiv b^{c*d} \equiv b^{c*d~(mod~J(m))} \equiv b^1 \equiv b {\rm ~(mod~} m)
$$
the {\em $c$-th root} $b$ of $a$. \par

@@

\begin{theorem}\label{thm-zth-ordp}\footnote{%
    For prime moduli $p$ all $a$ with $0 < a < p$ are of order $J(p) = p - 1$.
    Compare table~\ref{expmod45} for an example. In this case $a^i (\mbox{mod }n)$ goes
    through all the values $1,\dots,p-1$. Exhausting all possible values of
    the set is an important cryptographic proposition (compare theorem
    \ref{thm-zth-exhperm}). This determines a permutation\index{Permutation}
    $\pi(p-1)$.}$^,$\footnote{%
    Table~\ref{modexp46} demonstrates that for composite moduli $n$ not all $a$ are of
    maximal order $J(n)$. In this example only $5,7,11,15,17,19\mbox{ and }21$
    are of order 22.
  } Given a modulus $n$ and a number $a$, relative prime to $n$ the
  following holds: \\
  The set $\{ a^i\; (\mbox{mod }n) |\; i = 1,\dots,J(n)\}$ equals the
  multiplicative group $Z_n^*$ if and only if $ord_n(a) = J(n)$.
\end{theorem}
The multiplicative group $Z_n^*$ contains all values from $1$ to $n-1$.

@@

% --------------------------------------------------------------------------
\hypertarget{RSA}{}
\subsection{How the RSA procedure works}
\label{RSA}
The individual steps for implementing the \index{RSA} RSA procedure can be described as follows (see \cite[p. 213 ff]{nt:Eckert2003}\index{Eckert 2003} and 
\cite[p. 338 ff]{nt:Sedgewick1990}\index{Sedgewick 1990}). 
Steps 1 to 3 constitute key generation, steps 4 and 5 are the encryption, and steps 
6 and 7 are the decryption:
\begin{itemize}
\item[{\bf 1.}] Select two distinct random prime numbers\footnote{
Compaq introduced the so-called multi-prime method with high marketing 
effort in 2000.
$n$ was the product of two big and one relative small prime: $n=o*p*q.$ 
With theorem~\ref{J_of_p1..pk} we get:
$J(n)= (o-1)*(p-1)*(q-1). $ This method did not assert itself yet.\\
One reason probably is, that Compaq claimed a patent \index{Patent} on it. 
Generally there is less understanding in Europe and with the Open Source
Initiative\index{Open Source}, that one can claim patents on algorithms.
But there is really no understanding outside the U.S., that one can get
a patent for a special case (3 factors) of an algorithm (RSA), although
the patent for the general case was almost expired.}$^,$\footnote{If the
two primes $p$ and $q$ are equal then $ (m^{e})^{d} \equiv m $ mod $n$ is
not true for all $ m<n$ (although $ e*d \equiv 1$ mod $J(n)$ is fulfilled). 
\begin{example}{:}\\
If $ n=5^2$ then according to theorem~\ref{thm-zth-phinum} it 
is $J(n)=5*4=20, ~ e=3, ~ d=7, ~e*d=21\equiv 1$ mod $J(n).$
\end{example}
But it is $ (5^3)^7\equiv 0 $ mod $ 25$. } $p$ and $q$ and calculate 
$n = p*q$\footnote{The GISA\index{GISA} (German Information Security Agency) 
recommends, to choose the prime factors $p$ and $q$ almost the same,
but not too close:
$$ 0.5 < |\log_2 (p) - \log_2 (q) | <30. $$
They recommend to generate the primes independently and check that the 
restriction is fulfilled (see \cite{nt:GISA2002}).
}.\\
The value $n$ is called the RSA modulus\footnote{
In CrypTool\index{CrypTool} the RSA modulo is denoted with a capital ``$N$'' .}.

@@

\noindent We will show this in 3 steps (see \cite[p. 131ff]{nt:Beutelspacher1996}).

\noindent {\bf Step 1}:

@@

\noindent {\bf Step 2}:

@@

\noindent {\bf Step 3}:

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{SecurityRSA}{}
\section[Considerations regarding the security of the RSA algorithm]
{Considerations regarding the security of the RSA algorithm\footnotemark}
    \footnotetext{Major
    parts of chapters~\ref{complexity} and \ref{chptSecurityParam} follow
    the article ``Vorz\"uge und Grenzen des RSA-Verfahrens'' written by
    F.~Bourseau, D.~Fox and C.~Thiel \cite{nt:Bourseau2002}.}
    \label{SecurityRSA}

@@

The best factorization method known today is a further development of the
General Number Field Sieve (GNFS) \index{General Number Field Sieve (GNFS)},
which was originally devised to factor only numbers of a special form 
(like Fermat numbers).  The complexity of solving the factorization problem
with the GNFS is asymptotically
$$
O(l) = e^{c \cdot (l \cdot \ln 2)^{1/3} \cdot  (\ln(l \cdot \ln(2))^{2/3} + o(l)}
$$
Please refer to: 
\vspace{-10pt}
\begin{itemize}{}
  \item A. Lenstra, H. Lenstra:  
          {\em The development of the Number Field Sieve} 
          \cite{nt:Lenstra1993}.
  \item   Robert D. Silverman:  
          {\em A Cost-Based Security Analysis of Symmetric and Asymmetric
          Key Lengths} 
          \cite{nt:Silverman2000}.
\end{itemize}

@@

Please refer to: {\em Handbook of Applied Cryptography} \cite{nt:Menezes2001}.

@@

      Please refer to:
      \begin{list}{}{\setlength{\topsep}{-7 pt}}
        \item[] C. Pomerance:  
                {\em The quadratic sieve factoring algorithm} 
                \cite{nt:Pomerance1984}.
      \end{list}
      \vskip +12pt

@@

      Please refer to:
      \begin{list}{}{\setlength{\topsep}{-7 pt}}
        \item[] J.P. Buhler, H.W. Lenstra, C. Pomerance:  
                {\em Factoring integers with the number field sieve} 
                \cite{nt:Buhler1993}.
      \end{list}
      \vskip +12pt
\end{itemize}

@@

A good overview about the current status of publications concerning these attacks
can be found in the diploma thesis of Matthias Schneider \cite{nt:SchneiderM2004}.

@@

An estimation of the evolution of secure key lengths was done by 
Lenstra/Verheul\cite{nt:Lenstra1999} (compare figure~\ref{RSAKeylength}
in chapter ~\ref{ECAlternative}).

Within the article \cite{nt:Bourseau2002} Dirk Fox\footnote{%
His company Secorvo Ltd also delivered a statement on the recommendation for key
length selection published by the GISA\index{GISA} (German Information
Security Agency). Chapter 2.3.1 of this statement contains a competent
and understandable discussion of RSA security (this document
exists -- to my knowledge -- only in German):\\
\href{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}
{\texttt{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}}
}
published his prognosis of an almost linear factorization progression,
if all influencing factors are included: Each year the module length feasible for
factorization increases by 20 bit on average. So his forecast was below the
more optimistic estimations of GISA and NIST.

@@

In his paper ``Circuits for integer factorization: a proposal'' 
(\href{http://cr.yp.to/djb.html}{\texttt{http://cr.yp.to/djb.html}}),
published November 2001, D.~J.\ Bernstein \cite{nt:Bernstein2001} 
addresses the problem of
factorizing large integers. Therefore, his results are of relevance from a
RSA point of view.  As a main result Bernstein claims that the
implementation of the General Number Field Sieve algorithm (GNFS)
 \index{General Number Field Sieve (GNFS)} can be improved to factor, with
the same effort as before, integers with three times more digits.

@@

Arjen Lenstra, Adi Shamir et. al. analyzed the paper of Bernstein
\cite{nt:Lenstra2002}.  In summary they expect a factorization improvement on
how much longer the bit length of the keys could be with a factor of 1.17
(instead of factor 3 as proposed by Bernstein).

@@

\href{http://www.rsasecurity.com/}{RSA Security's} analysis of the Bernstein
paper \cite{nt:RSA Security 2002} from April, 8 2002 also -- as expected --
concludes, that RSA is still not compromised.

@@

In January 2003 Adi Shamir and Eran Tromer from the Weizmann Institute of Science published a preliminary draft called {\em ``Factoring Large Numbers with the TWIRL Device''} raising concerns about the security of key sizes till 1024 bits \cite{nt:Shamir2003}.

@@

A detailed explanation from these two authors also can be found in the
RSA Laboratories CryptoBytes \cite{nt:Shamir2003a}.

The 3-page article in the DuD issue of June 2003 \cite{nt:Weis2003} contains
a very good explanation, how the attack using the Generalized Number Field
Sieve (GNFS) \index{General Number Field Sieve (GNFS)} works and which 
progress is made, to factorize numbers.
At GNFS we can distinguish 2 general steps: 
The sieve step (relation collecting) and the matrix reduction.
Besides the sieve step is highly parallelizable, it dominates the overall
calculation burden. Shamir and Tromer haven't built a TWIRL device yet,
but the estimated costs of 10 till 50 million Euro (in order to factorize
a 1024-bit number) is not prohibitive for secret agencies or big criminal
organizations, because the ``costs for a single espionage satellite is
estimated e.g. to be several billion USD''. The authors therefore
recommend, to get as soon as possible rid of today used sensible RSA, 
Diffie-Hellman or ElGamal keys up to 1024 bit and to use then keys of at
least 2048 bit length.
The planned TCPA/Palladium hardware \index{Palladium} will use 2048-bit
RSA keys!

@@

In August 2002 the three Indian researchers M. Agrawal, N. Kayal and N. Saxena published the paper {\em ``PRIMES in P''} about a new primality testing algorithm called AKS\index{AKS} \cite{nt:Agrawal2002}. 
They discovered a polynomial\index{Polynomial} time deterministic algorithm for determining if a number is prime or not.

@@

Alice and Bob\footnote{Bob\index{Bob} and Alice\index{Alice} are the default
names used for the two authorized participants in a protocol (see
\cite[p. 23]{nt:Schneier1996nt}).
} use a one way function to obtain a key $S$, the session key, for subsequent
correspondence. This is then a secret that is only known to the two of them.
Alice selects a random number $a$ and keeps it secret. She applies a one way
function to $a$ to calculate the number $A = g^a$ and sends it to Bob. He does
the same, by selecting a secret random number $b$, calculating $B = g^b$ and
sending it to Alice. The number $g$ is random and can be publicly known.
Alice applies the one way function together with her secret number $a$ to
$B$, while Bob does the same with his secret number $b$ and the received
number $A$.

@@

\noindent {\bf Key generation (steps 1 to 3)}:\\
{\bf 1.} $p=47, q=79$ $( n= 3713; ~J(n) = (p-1)*(q-1)=3588).$ \\
{\bf 2.} $e = 37$ (should\footnote{%
                See footnote~\ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $79$ and $3587$ and must be relatively prime
                to $3588$). \\
{\bf 3.} $d=97$ (since $e*d=1{\rm ~mod~}J(n); 37*97 \equiv 3589 \equiv 1{\rm ~(mod~}3588) \;$)\footnote{How
to compute $d = 97$ using the {\em extended} gcd algorithm is shown in \hyperlink{NumberTheory_Appendix_GCD}{appendix A of this chapter}}. \\

@@

\noindent The idea for this task\footnote{Using CrypTool\index{CrypTool} you can 
solve this via the menu path
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.} 
is taken from the example in \cite[p. 271]{nt:Eckert2003}\index{Eckert 2003}.

@@

The task is taken from \cite[Exercise 4.6]{nt:Stinson1995}\index{Stinson 1995}:
The pure solution has been published by Prof. Stinson at 
\href{http://www.cacr.math.uwaterloo.ca/~dstinson/solns.html}{\texttt{http://www.cacr.math.uwaterloo.ca/\~{}dstinson/solns.html}}.\footnote{or
\href{http://bibd.unl/~stinson/solns.html}{\texttt{http://bibd.unl/\~{}stinson/solns.html}}.}

@@

The following task is a corrected version from the book written by Prof. Yan 
\cite[Example 3.3.7, p. 318]{nt:Yan2000}\index{Yan 2000}.
However, it is not the result that is important here but rather the individual steps of the solution, that is, 
the explanation of the cryptanalysis\index{Cryptanalysis}\footnote{The method of solving the problem is outlined in the scenario of 
the online help to CrypTool\index{CrypTool} and in the CrypTool presentation. If anyone sends us a well prepared exact 
method of solving the problem, we would be pleased to include it in the documentation.}.

@@

\bibitem[Agrawal2002]{nt:Agrawal2002}  \index{Agrawal 2002} 
    M. Agrawal, N. Kayal, N. Saxena, \\
    {\em PRIMES in P}, August 2002, revised paper: \\
       \href{http://www.cse.iitk.ac.in/~manindra/algebra/primality_v6.pdf}
    {\texttt{http://www.cse.iitk.ac.in/~manindra/algebra/primality\_v6.pdf}} \\
    See also the website "The AKS "PRIMES in P" Algorithm Resource":
       \href{http://fatphil.org/maths/AKS/}
    {\texttt{http://fatphil.org/maths/AKS/}}

\bibitem[Bartholome1996]{nt:3Bartholome1996}  \index{Bartholome 1996} 
    A. Bartholome, J. Rung, H. Kern, \\
    {\em Zahlentheorie f\"ur Einsteiger}, Vieweg 1995, 2nd edition 1996.
